### 2017-01-22 draft-manuscript.Rmd
added new content and bibliography

---
title             : "A Computational Perspective On Spatial Perspective Transformations: From Sensory Inference To Imagined Self-motion"
# title             : "Probabilistic Computations Underlying Simulated Self-Rotations"
shorttitle        : "A Computational Perspective"

author: 
  - name          : "Andrew W. Ellis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Fabrikstrasse 8, 3012 Bern, Switzerland"
    email         : "andrew.ellis@psy.unibe.ch"
  - name          : "Fred W. Mast"
    affiliation   : "1"

@@

bibliography      : ["references.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
# output            : papaja::apa6_word
output            : papaja::apa6_pdf
---

@@

@@

# Introduction

@@

@@

![SPT and imagined self-motion involve common computations](diagrams/fig-1-common-process.pdf)
But what are these computations? The brain could either solve the equation for the final position, or could run foward simulation of the dynamic system.

@@

## Imagined Self-Motion

@@

@@

>>>>

---
    title             : "A Computational Perspective On Spatial Perspective Transformations: From Sensory Inference To Imagined Self-motion"
# title             : "Probabilistic Computations Underlying Simulated Self-Rotations"
# title: Probabilistic Computations Underlying Vestibular Cognition
shorttitle        : "A Computational Perspective"

author:
    - name          : "Andrew W. Ellis"
# affiliation   : "1"
corresponding : yes    # Define only one corresponding author
address       : "Fabrikstrasse 8, 3012 Bern, Switzerland"
email         : "andrew.ellis@psy.unibe.ch"
- name          : "Fred W. Mast"
# affiliation   : "1"

@@

bibliography      : ["bibliography.bib"]

figsintext        : yes
figurelist        : yes
tablelist         : yes
footnotelist      : no
lineno            : yes
citeproc          : yes
csl               : apa.csl

lang              : "english"
class             : "man"
output            : papaja::apa6_word
# output            : papaja::apa6_pdf
---

@@

> Perspective taking: ability to project our body image onto other actors.

- higher level variables generate an entire sequence of control commands
- show models:
    + passive model for low-level sensory inference
+ active model with control input for anticipating sensory consequences of active movement
+ the active model can be used for:
    - estimating ex-afference
- "interpreting" senosry signals when these are very noisy/difficult. In other words, when the stimulus is difficult, we need to interpret the data. We claim that this makes use of the control input model.
- intereference as the result of
+ inappropriately applying active control model for sensory estimation, leadin to mis-specified model
+ wrong prior during model selection
- show various scenarios for interference:
    + decision making for motion direction discrimination
+ imaginining motion may result in inadvertently using an active motion model for inference in a subsequent trial. IN a discrimination trial, participants need to perfom decision-making. THerefore, we need to understand decision-making. We present a model of how decision-making in a 2afc experiment may occur (Bayesian model selection -> sequential Bayes factor)
+ if imagery was in a specific direction, this may result in an expectation. Define prior, expectation, anticipation, etc.

@@

@@

Why does the brain need to simulate? Because in the case of complex, non-linear transformations, the brain does not have acces to analytical solutions, and must therefore rely on simulations. Intuitively, the brain uses the 'cheap' solution of simulating a dyanmic system (i.e. movement of the body through space), because it already posseses the machinery to perform these simulations, and because analytic solutions to complex mulit-dimensional rotations and translation are hard to obtain.

@@

![SPT and imagined self-motion involve common computations](diagrams/fig-1-common-process.pdf)

But what are these computations? The brain could either solve the equation for the final position, or could run foward simulation of the dynamic system.

> SPT and mental imagery as active construction, not just memory recall

@@

## Imagined Self-Motion
[@Tian:2012ui]
Is mental imagery an internal predictive process?

Mental imagery is further hypothesized to be a predictive process (for future perceptual states), in which the dynamics of perceptual experience can be retrieved/calculated and reconstructed internally (Moulton and Kosslyn, 2009).

Mental imagery is oftern portrayed as being the result of top-down reconstruction of activity in perceptual circuits in the absence of physical stimulation of the sensors, and is opposed to bottom-up perception. However, in the case of the vestibular system, it becomes clear that this view is inadequate (the same is true for other modalities): the sensory that the vestibular sensors provide is noisy and ambiguous (examples), and in order to deal with these facts, the brain must use several strategies (which are well-known from control theory). It is assumed that the brain uses (at least approximatively) statistically optimal filtering algorithms. Initia research cast the type of processing in the framework of control theory (optimal observer models); however @Selva:2012gsa showed that these models are equivalent to statistically optimal models (Kalman filters), in the case of linear dynamical systems.

That is, given an initial point, the series of future perceptual states can be internally simulated by following the regularity (temporal and causal constraints) stored in declarative memory (general knowl- edge). The mapping between internal simulation and the per- ception of external stimulation is thought not to be necessarily isomorphic (Goldman, 1989), as only the essential intermediate states are required to have a one-to-one mapping (Fisher, 2006).

@@

# Interferene
We walk through several common conceptions of "interference" and explain these in terms of an agent performing mental simulations whilst simultaneously attempting to interact with its environment by performing sensory inference.

We can define interference as the phenomenon that sensory inference (filtering) is affected by another process (simulation) using (parts of) the same generative model. The agent has two goals: firstly, it has to interact with the environment, using its sensors, and it has to analyze the data. Secondly, the agent wishes to plan or simulate, using the same generative model. It is currently unclear to what extent the brain is able to segregate the two processes.

@@

# Particle Filtering
Using difference equations, the stochastic state space model can be written as:
    $$ x_{t} = f(x_{t-1}, u_{t-1}, \nu) $$
    $$ y_{t} = g(x_{t}, \eta) $$

    where ...

The state space model consists of an unobservale state process, $x$, and an observable process $y$. The latent state is modelled as a first-order Markov process, which means that $x_t$ is conditionally independent of the past, given the previous value. Therefore, the latent process is completely described by its iniial conditions, $x_0$, and the transition distribution. The observation process depends only on the current state.

### 2017-01-22 manuscript.Rmd
added new content and bibliography

---
title             : "A Computational Perspective On Spatial Perspective Transformations: From Sensory Inference To Imagined Self-motion"
shorttitle        : "A Computational Perspective"

author: 
  - name          : "Andrew W. Ellis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Fabrikstrasse 8, 3012 Bern, Switzerland"
    email         : "andrew.ellis@psy.unibe.ch"
  - name          : "Fred W. Mast"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Department of Psychology, University of Bern"

author_note: >
  Andrew W. Ellis, Fred W. Mast, Department of Psychology.

@@

  Raw data and analysis scripts are available at https://osf.io/u26mq.

abstract: >
    Recent research showed the involvement of vestibular information in cognitive processes. We provide a computational framework for vestibular cognition and thereby focus on perspective transformations, which require a simulated rotation of the self. We explore how the ability to model the effects of body rotation using vestibular input might enable the ability to imagine body rotations. We approach this problem by considering a simple agent endowed with a complex statistical model of the hidden causes of its sensory input. We explain how this statistical model enables the agent to predict its sensory input and hidden sensory variables. Based on a probabilistic graphical model, we formalize qualitative concepts of mental simulation. Predicting the sensory input is key to understanding imagined rotations. Mental imagery goes beyond prediction, rather, it involves the use of counterfactual queries in a probabilistic model.

keywords          : Probabilistic graphical models, Particle filter, Embodied cognition, Vestibular cognition, Spatial perspective taking, Mental imagery
wordcount         : "X"

bibliography      : ["references.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
# output            : papaja::apa6_word
output            : papaja::apa6_pdf
---

```{r message = FALSE, warning = FALSE}
library("papaja")
```

# Results

# Discussion

>>>>

---
title             : "Vestibular Contributions to Spatial Perspective Transformations: From Sensory Inference To Imagined Self-Motion"
# title             : "A Computational Perspective On Spatial Perspective Transformations: From Sensory Inference To Imagined Self-motion"
# title             : "Probabilistic Computations Underlying Simulated Self-Rotations"
# title: Probabilistic Computations Underlying Vestibular Cognition
shorttitle        : "Vestibular Contributions to SPT"

author:
  - name          : "Andrew W. Ellis"
    # affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Fabrikstrasse 8, 3012 Bern, Switzerland"
    email         : "andrew.ellis@psy.unibe.ch"
  - name          : "Fred W. Mast"
    # affiliation   : "1"

affiliation:
  # - id            : "1"
  - institution   : "Department of Psychology, University of Bern, Switzerland"

author_note: >
  Andrew W. Ellis & Fred W. Mast, Department of Psychology.

@@

  Code for simulations is available at https://osf.io/u26mq.

abstract: >
    Recent research has indicated that the vestibular network is involved in cognitive processes. However, the nature of this involvement remains unclear. We provide a computational framework for vestibular cognition and discuss this in the context of perspective transformations, which require a simulated rotation of the self. We explore how the ability to model the effects of body rotation using vestibular input might enable the ability to imagine body rotations. We approach this problem by considering a simple agent endowed with a complex statistical model of the hidden causes of its sensory input. We explain how this statistical model enables the agent to predict its sensory input and hidden sensory variables. Based on a probabilistic graphical model, we formalize qualitative concepts of mental simulation. Predicting the sensory input is key to understanding imagined rotations. Mental imagery goes beyond prediction, rather, it involves the use of counterfactual queries in a probabilistic model.

keywords: probabilistic graphical models, particle filter, embodied cognition, vestibular cognition, spatial perspective taking, mental imagery
wordcount: "X"

bibliography: "bibliography.bib"

figsintext        : yes
figurelist        : yes
tablelist         : yes
footnotelist      : no
lineno            : yes
citeproc          : yes
csl               : apa.csl

lang              : "english"
class             : "man"
# output            : papaja::apa6_word
output:
     papaja::apa6_pdf:
        pandoc_args:
        - --filter
        - pandoc-fignos
fignos-cleveref: On
fignos-plus-name: Figure
header-includes: \usepackage{caption}
---

```{r preliminary, message = FALSE, warning = FALSE}
library(papaja)
library(pander)
library(ggplot2)
library(dplyr)
library(tidyr)
library(mvtnorm)
# library(ggthemes)
library(cowplot)

source('~/thesis/papers/pgm-spt/R/utils.R')
source('~/thesis/papers/pgm-spt/R/2d-particle-filter.R')
source('~/thesis/papers/pgm-spt/simulations/rbiips-model.R')

knitr::opts_chunk$set(cache = TRUE,
                      fig.width = 12, fig.height = 8,
                      dpi = 600, fig.path = "../generated-figures/",
                      dev=c('pdf', 'png', 'cairo_ps'))
```

# Introduction
 <!-- convey how the work constraints current theorizing over and above the current literature -->

<!-- The aim of this paper is to describe potential contributions of the vestibular system to spatial perspective transformations in the context. of probabilistic sensory inference. -->

Adopting the spatial perspective of another person is generally considered an essential cognitive ability, and plays a vital role in our ability to determine what another person can see and to understand another person's action possibilities, in order to coordinate joint actions [@CreemRegehr:2013kx, @Pezzulo:2013iy]. Many studies have also pointed out that is of fundamental importance for language comprehension and communication [@Beveridge:2013gx], and social cognition in general [@Deroualle:2014ho].

A number of recent studies have demonstrated that the vestibular system, which deals with sensory signal relating to movements of the head, is involved in such spatial transformations [@Lenggenhager:2008et, @Deroualle:2015gk, @Falconer:2012cl, @Gardner:2016kd]. However, to date, there has little attempt to understand the involvement of the vestibular system in higher-level cognitive abilities from a computational perspective. In this paper, we focus specifically on the role of the vestibular system for spatial perspective transformations, and we draw the link between probabilistic computations performed by the vestibular system in the service of real-time interaction with the world, and the brain's ability to run mental simulations for the purpose of adopting the perspective of another person.

Consider the use of deictic pronouns, such as "this" or "that" and "left" or "right". These terms only assume a their meaning when placed within a certain reference frame. For example, in a dialogue between two speakers, the statement "I would like that one", or "I would like the one on the right" can only be understood under the assumption that the deictic terms "that" and "right" are used in an egocentric reference frame. In addition, both interlocutors must be able to access the other's reference frame with respect to an external reference frame, which is independent of both interlocutors. In other words: In order to give instructions, the speaker must be able to mentally assume the addressee's position. A further example that requires a similar spatial transformation is when two people want to perform a coordinated action, such as lifting a heavy object. This requires having knowledge of the other's potential actions in order to adjust one's own actions, and this again requires knowledge of the other's position with an external reference frame. Furthermore, it is considered likely that merely possessing a visual representation is not sufficient [citation]. A judgement about what kind of actions the other person can perform, requires an embodied simulation using a representation of one's own body schema [@Pezzulo:2013iy]. The information necessary to perform this simulation is thus the position of the addressee's limbs, in their egocentric body coordinates, the position of the addressee's reference frame realtive to the external reference frame, and the difference between the speaker and the addressee's egocentric reference frames. Perspective taking thus can be conceived of as a aligning a representation of one's body with that of another person in order to access spatial information from a viewpoint other than one's own egocentric viewpoint. In order to achieve this alignment, one must perform translations and rotations of one's own egocentric reference frame. Processing translations and rotations of the head and body is the core domain of the vestibular system, and thus is of particular interest for research into mental spatial transformations. Spatial transformations in 6 dimensions (translation along 3 axes, and rotations around 3 axes) are highly complex, and non-linear, and it is commonly assumed that such transformations of perspective are performed by running embodied simulations [@Pezzulo:2013iy]. Thus, it seems that the brain adopts the strategy of simulating motion of the body through space in order achieve a desired target position. This effectively amounts to a running a simulation of dynamical system, consisting of limb movements and the resulting sensory consequences. Such a simulation has therefore been hypothesized to involve both motor and sensory components [citation]. @fig:spt-schematic (a) shows a simple scenario in which a speaker wants to give an addressee, standing opposite, verbal instructions regarding the location of an external object. In order to give appropriate instructions, the speaker must determine the most suitable action, given the addressee's position relative to the external object, and in order to achieve this goal, the speaker must align himself with the addressee, so that he can use his own body-centred reference frame to quickly simulate the best action. In this example, alignment with the addressee's reference frame consists of a forwards translation and a rotation around the vertical axis. This can be described as an imagined movement [citation], in which the goal is achieve a specific position. In order to do this, the speaker must first determine the parameters related to the movement, such as the type of movement, and the direction, velocity and duration that lead to the desired outcome. This means that in order to know what to simulate, the speaker must work backwards from the desired outcome, and infer the actions that are most likely to lead to that outcome. This is a well-known problem in the field of reinforcement learning, and planning in humans has been described in similar terms, as reverse inference [@Botvinick:2012fe]. @fig:spt-schematic (b) shows that if the goal is to simply simulate the perceptual experience of rotation, rather than project a representation of one's body onto a target, the problem is vastly simplified; however, the computational steps remain essential the same. In order to imagine a whole-body rotation around the vertical axis, the same parameters must be chosen. In this study, we focus on the simplified problem of imagined rotation around the body's vertical (yaw) axis.

![SPT and imagined self-motion involve common spatial transformations of a representation of the self. (a) If a speaker's goal is to take the perspective of an addressee, he must perform a translation and rotation of his own egocentric reference frame. These operations can be viewed as imagined movements.(b) Focussing on just the rotation around the vertical axis allows us to simplify the problem, and highlights the fact that perspective taking and mental imagery involve common computations.](diagrams/fig-1-common-process.pdf){#fig:spt-schematic}

Performing a simulation of the behaviour of a dynamical system requires an internal model of the system. Whereas earlier notions of internal models followed the control-theoretic approach taken in sensorimotor research [@Wolpert:1998cc, @Shadmehr:2012vu], this idea has recently been applied more generally due to the use of Bayesian models of perception and cognition [Friston, Griffiths, Tenenbaum].

[@Pickering:2014bo] discuss the relationship between control-theoretic notions of forward models and and more genereal concepts of a generative model in a hierarchical Bayesian framework.

In short, running a simulation of a system requires a generative model. Where does this knowledge come from? __Here we make the link to models of vestibular sensry processing__.

For this reason, we claim that the vestibular system offers a very promising field of research. On the one hand,

## Involvement of the vestibular system
If the vestibular system is involved in spatial perspective taking, then this must be related to mental whole-body rotation, as the vestibular system is directly concerned with rotations and translations. In particular, the semi-circular canals (SCC) measure angular accelerations, and the upstream vestibular system provides the neural circuitry involved in processing these sensory measurements. The vestibular system faces a number of challenges in interpreting these sensory measurements. On the one hand, the signal provided by the SCC and otoliths are very noisy, and ambiguous [MacNeilage]; the otoliths respond to translations and tilt of the head relative to gravity, and in order to disambiguate between translation and tilt, the brain must perform sophisticated computations at the level of the brain stem. It has become clear that the nature of these computations is probabilistic, and that the brain makes use of strong prior beliefs in order to interpret these sensory signals.

A further challenge faced by the brain is that it cannot determine whether the sensory signals provided by the SCC and otolith signals m

## Focus on vestibular system
- active motion, efference copy, etc
- generative model

- mental simulations involves forward models
- interesting, because vestibular system is well-described in terms of control theory

<!-- give a few short examples: -->
<!-- - deictic pronouns: here, there, left, right, this or that -->
<!-- - performing joint actions (shared action space): lifting a heavy object -->

Imagined motion through space vs. rotating contents of visual buffer

# Dynamic Bayesian models for sensory inference

## Related Work: Previous models for sensory inference
We begin by briefly reviewing previous probabilistic models of sensory inference:
Whereas early studies largely adopted approached inspired by control theory, several recent studies have described vestibular sensory processing as dynamic Bayesian inference [@PaulRMacNeilage:2008bv, @Karmali:2012cv, @Laurens:2006be]. In particular, @Karmali:2012cv and @Laurens:2006be described sensory inference in terms of particle filtering, which is a particular inference algorithm that can be used to perform inference in a non-linear state-space model [@Doucet:2000bh, @Doucet:2009us].

State-space models can be seen as special cases of more general Bayesian hierarchcal models [@Bishop:2006ui; @Murphy:2012ua; @Koller:2009ty].

## Summary

## Reframe as PGM

# Extend previous models with control input

Here we extend the basic particle filtering (algorithm) model, in order to:

- provide the capability to model the sensory consequences of self-initiated actions
- provide the capability to run a Monte Carlo simulation using the same generative model as is used for sensory inference

## State-Space Model

## Inference Algorithm 
Sequential Monte Carlo

# Modelling questions
Experimental results to be explained: @Wertheim:2001cb report a cognitive suppression of otolith responses, based on participants' prior beliefs about possible motions.

# An Investigation of Various Priors
Here, we identify the various forms of predictive processing in the context of real-time sensory inference using SMC
## Related to Filtering Algorithm
## Higher-level Priors

# Various sources of interference
Here, we identify various sources of intereference (2-way) between thought (higher-level cognition) and lower-level sensory processing

## Model mis-specification
## Re-use of same generative model
This may lead to leakage, i.e. inadvertently incorporating sensory evidence

## Decision-making
The interference may be explained in terms of higher-level perceptual decision -making mechanisms
## evidence accumulation
## bias (Starting point)

# Simulations

# General Discussion

# Summary and conclusions
# Motion Data

```{r generate-data-1, echo=FALSE, message=FALSE, warning=FALSE}
dt <- 0.1
motiondata <-  generate_data(T = 2, frequency = 0.5,
                             amplitude = 20,
                             sensor_sd = 2.0)
# plot_trajectories(motiondata = motiondata)

motiondata_hidden <- motiondata %>%
    gather(type, value = value, -time) %>%
    filter(type != "observations") %>%
    mutate(type = factor(type,
                         levels = c("position",
                                    "velocity",
                                    "acceleration")))

p1 <- motiondata_hidden %>%
        ggplot(aes(x = time, y = value)) +
            geom_line(aes(linetype = type), size = 1.4) +
            geom_point(data = motiondata,
                       aes(x = time, y = observations),
                       shape = 21, colour = "white",
                       fill = "white", size = 6, stroke = 2) +
            geom_point(data = motiondata,
                       aes(x = time, y = observations),
                       shape = 21, colour = "black",
                       fill = "white", size = 4, stroke = 2) +
            guides(linetype = guide_legend(keywidth = 2.5, keyheight = 1)) +
            xlab("Time (sec)") +
            ylab("Angular position (deg)") +
            theme_apa() +
            theme(legend.title=element_blank())

motiondata_2 <-  generate_data(T = 2, frequency = 0.5,
                             amplitude = -20,
                             sensor_sd = 1.8)
# plot_trajectories(motiondata = motiondata)

motiondata_hidden_2 <- motiondata_2 %>%
    gather(type, value = value, -time) %>%
    filter(type != "observations") %>%
    mutate(type = factor(type,
                         levels = c("position",
                                    "velocity",
                                    "acceleration")))

p2 <- motiondata_hidden_2 %>%
        ggplot(aes(x = time, y = value)) +
            geom_line(aes(linetype = type), size = 1.4) +
            geom_point(data = motiondata_2,
                       aes(x = time, y = observations),
                       shape = 21, colour = "white",
                       fill = "white", size = 6, stroke = 2) +
            geom_point(data = motiondata_2,
                       aes(x = time, y = observations),
                       shape = 21, colour = "black",
                       fill = "white", size = 4, stroke = 2) +
            guides(linetype = guide_legend(keywidth = 2.5, keyheight = 1)) +
            xlab("Time (sec)") +
            ylab("Angular position (deg)") +
            theme_apa() +
            theme(legend.title=element_blank())

cowplot::plot_grid(p1, p2, nrow = 2, labels = c("a", "b"))
```

# Test Simulations

```{r passive, echo=FALSE, message=FALSE, warning=FALSE}
params <- list(sdx = c(0.8, 0.8) * eye(2), sdy = 2.0,
                x_init = c(0, 0),
                sdx_init = 0.5*eye(2),
                A = 20, N = 1000,
                dt = 0.1, freq = 0.5,
                transfun = transfun)

passive <- run_particle_filter(motiondata = motiondata, params = params,
                            resample_particles = TRUE, rs_thresh = 0.5,
                            selfmotion = "passive")

plot(passive, panel_label = "", print_loglik = TRUE, predict = FALSE, imagine = FALSE)
```

```{r echo=FALSE}
plot_pf_estimates(object = passive, predict = FALSE)
```

# Tables

  Right     Left     Center     Default
-------     ------ ----------   -------
     12     12        12            12
    123     123       123          123
      1     1          1             1

Table:  Demonstration of simple table syntax.

-------------------------------------------------------------
 Centered   Default           Right Left
  Header    Aligned         Aligned Aligned
----------- ------- --------------- -------------------------
   First    row                12.0 Example of a row that
                                    spans multiple lines.

  Second    row                 5.0 Here's another one. Note
                                    the blank line between
                                    rows.
-------------------------------------------------------------

Table: Here's the caption. It, too, may span
multiple lines.
